### 2025-07-23 paper/sections/experiments.qmd
add done

To assess outcomes with respect to actionability for non-linear models, we look at the average costs of valid counterfactuals in terms of their distances from factual starting points. While this an imperfect proxy of sensitivity, we hypothesize that CT can reduce these costs by teaching models to seek plausibility with respect to mutable features, much like we observe in @fig-poc in panel (d) compared to (c). We supplement this analysis with qualitative findings for integrated gradients [@sundararajan2017ig]. Finally, for predictive performance, we use standard metrics, such as robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].

>>>>

To assess outcomes with respect to actionability for non-linear models, we look at the costs of (just) valid counterfactuals in terms of their distances from factual starting points with $\tau=0.5$. While this an imperfect proxy of sensitivity, we hypothesize that CT can reduce these costs by teaching models to seek plausibility with respect to mutable features, much like we observe in @fig-poc in panel (d) compared to (c). We supplement this analysis with estimates for integrated gradients [@sundararajan2017ig]. Finally, for predictive performance, we use standard metrics, such as robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].

### 2025-07-23 paper/sections/other/counterfactuals.qmd
add done

In this section, we provide some background on gradient-based counterfactual generators (@sec-app-ce-background) and discuss how we define convergence in this context (@sec-app-conv).

@@

Gradient-based counterfactual search was originally proposed by @wachter2017counterfactual. It generally solves the following unconstrained objective,

$$
\begin{aligned}
\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(g(\mathbf{z}^\prime)) }  \right\} 
\end{aligned} 
$$

where $g: \mathcal{Z} \mapsto \mathcal{X}$ is an invertible function that maps from the $L$-dimensional counterfactual state space to the feature space and $\text{cost}(\cdot)$ denotes one or more penalties that are used to induce certain properties of the counterfactual outcome. As above, $\mathbf{y}^+$ denotes the target output and $\mathbf{M}_{\theta}(\mathbf{x})$ returns the logit predictions of the underlying classifier for $\mathbf{x}=g(\mathbf{z})$.

For all generators used in this work we use standard logit crossentropy loss for $\text{yloss}(\cdot)$. All generators also penalize the distance ($\ell_1$-norm) of counterfactuals from their original factual state. For *Generic* and *ECCCo*, we have $\mathcal{Z}:=\mathcal{X}$ and $g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}$, that is counterfactual are searched directly in the feature space. Conversely, *REVISE* traverses the latent space of a variational autoencoder (VAE) fitted to the training data, where $g(\cdot)$ corresponds to the decoder [@joshi2019realistic]. In addition to the distance penalty, *ECCCo* uses an additional penalty component that regularizes the energy associated with the counterfactual, $\mathbf{x}^\prime$ [@altmeyer2024faithful].

<!-- TODO: Add note on why we omit the conformal prediction component. -->

>>>>

In this section, we provide some additional background on gradient-based counterfactual generators (@sec-app-ce-background) and discuss how we define convergence in this context (@sec-app-conv).

@@

Gradient-based counterfactual search was originally proposed by @wachter2017counterfactual. A more general version of the counterfactual search objective presented in the paper is as follows,

$$
\begin{aligned}
\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{reg}(g(\mathbf{z}^\prime)) }  \right\} 
\end{aligned} 
$$

where $g: \mathcal{Z} \mapsto \mathcal{X}$ is an invertible function that maps from the $L$-dimensional counterfactual state space to the feature space and $\text{reg}(\cdot)$ denotes one or more penalties that are used to induce certain properties of the counterfactual outcome. As above, $\mathbf{y}^+$ denotes the target output and $\mathbf{M}_{\theta}(\mathbf{x})$ returns the logit predictions of the underlying classifier for $\mathbf{x}=g(\mathbf{z})$.

For all generators used in this work we use standard logit crossentropy loss for $\text{yloss}(\cdot)$. All generators also penalize the distance ($\ell_1$-norm) of counterfactuals from their original factual state. For *Generic* and *ECCCo*, we have $\mathcal{Z}:=\mathcal{X}$ and $g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}$, that is counterfactual are searched directly in the feature space. Conversely, *REVISE* traverses the latent space of a variational autoencoder (VAE) fitted to the training data, where $g(\cdot)$ corresponds to the decoder [@joshi2019realistic]. In addition to the distance penalty, *ECCCo* uses a penalty that regularizes the energy associated with the counterfactual, $\mathbf{x}^\prime$ [@altmeyer2024faithful]. We omit the conformal set size penalty proposed in the original paper, since a) the authors found faithfulness to primarily depend on the energy penalty and hence this alleviates us from one additional hyperparameter.

### 2025-07-23 paper/sections/other/main_results.qmd
add done

As discussed the main paper, CT is sensitive to certain hyperparameter choices. We study the effect of many hyperparameters extensively in @sec-app-grid. For the main results, we tune a small set of key hyperparameters (@sec-app-tune). The final choices for the main results are presented for each data set in @tbl-final-params along with training, test and batch sizes.

@@

## Final Results {#sec-app-final-results}

Plus/minus two standard deviations of bootstrap estimates.

<!-- Standard deviation of bootstrap is the standard error. -->

@@

### Robust Performance Plots

@@

@@

Mean outcomes for **CT** and **BL** along with bootstrapped confidence intervals (99%) for difference in mean outcomes grouped by dataset and evaluation metric. Column **LB** and **UB** show the lower and upper bound of the intervals, respectively, and computed using the percentile method. The underlying counterfactual evaluations are the same as the ones used to produce the main table in the paper.

@@

::: {#tbl-panel layout-ncol=3}

@@

Costs

@@

Validity

@@

Validity

>>>>

As discussed the main paper, CT is sensitive to certain hyperparameter choices. We study the effect of many hyperparameters extensively in @sec-app-grid of this appendix. For the main results, we tune a small set of key hyperparameters (@sec-app-tune). The final choices for the main results are presented for each data set in @tbl-final-params along with training, test and batch sizes.

@@

<!-- NOTE: Computing final results below -->

@@

<!-- NOTE: Generating robust performance plot below -->

@@

@tbl-ci present the exact confidence intervals (99%) for the difference in mean outcomes on which we base our assessment of statistical significance in the main paper. Grouped by evaluation metrics (Variable) and dataset (Data), the table presents the mean outcomes for CT and BT and finally the lower bound (LB) and upper bound (UB) of the confidence interval. To compute the intervals, we used the percentile method for bootstrapped confidence intervals: the lower and upper bound represent the $\alpha/2$- and $(1-\alpha / 2)-$quantile of the bootstrap distribution, respectively, for $\alpha=0.01$.

@@

Mean outcomes for CT and BL along with bootstrapped confidence intervals (99%) for difference in mean outcomes grouped by dataset and evaluation metric. Column LB and UB show the lower and upper bound of the intervals, respectively, and computed using the percentile method. The underlying counterfactual evaluations are the same as the ones used to produce the main table in the paper.

@@

In @tbl-panel, we present additional outcomes for common evaluation metrics: @tbl-costs presents the average reduction in costs of counterfactuals for CT vs. BL with no mutability constraints, i.e. corresponding to the first two columns in the main table of the paper; @tbl-val shows the corresponding average validities; finally, @tbl-val-mtbl shows average validities for the case with mutability constraints, i.e. corresponding to the third columns in the main table of the paper.

As noted in the discussion section of the main paper, we observe mixed results results here. Average costs in terms of distances from factual values decrease for most datasets, which is positively surprising since improved plausibility requires counterfactuals to travel further into the target domain than minimum distance counterfactuals. It appears that in these cases faithful counterfactuals for the baseline model still end up far away from their initial starting points, but not close enough for samples in the target domain to be plausible. In that sense, CT can be seen to improve both plausibility and costs for faithful CE. In some cases though (*LS*, *CH*, *MNIST*), we do seem to observe the tradeoff between plausibility and costs play out, as we would expect (compare panels (a) and (b) of Figure 1 in the main paper for reference).

Concerning validity, we find that can lead to substantial reductions and only increases average validity compared to the baseline in one case (*Circ*). As noted in the discussion section of the main paper, this result does not surprise us: by design, CT shrinks the solution space for valid counterfactual explanations, thus making it "harder" to reach validity compared to the baseline model. Note that for a number of reasons this should not be seen as problematic:

1. Validity of gradient-based CE is a function on the number of steps and the step size which we both kept fixed during evaluation: simply adjusting $T=50$ to higher values or choosing a larger step size will lead to higher rates of validity.
2. Even though reaching validity is sometimes "harder" in terms of the necessary number of steps for a given step size, we have already shown that the average distances that counterfactuals need to travel decrease for most datasets. Users care about costs in terms of feature distances, not search iteration steps.
3. From a philosophical perspective on algorithmic recourse, validity in and off itself is not a sufficient desideratum for counterfactuals. In fact, @ven2020philosophical propose introducing an upper bound on costs of the flipset (i.e. the set of valid CE), arguing that valid but highly costly counterfactuals are not useful to individuals in practice. In a similar fashion, it could be argued that there should be an upper bound on the implausibility of counterfactuals in the flipset.

::: {#tbl-panel layout="[[40,-20,40], [100]]"}

@@

Reduction in average costs for CT vs. the baseline. Results correspond to the case with no mutability constraints in the main table of the paper.

@@

Average validities of counterfactuals for CT and BL. Unconstrained case.

@@

Average validities of counterfactuals for CT and BL. Mutability constrained case.

### 2025-07-23 paper/sections/other/training.qmd
add done

@nte-eval-default presents the default hyperparameters used during evaluation.

@@

dict = Logging.with_logger(Logging.NullLogger()) do  
    CTExperiments.to_dict(EvaluationConfig(ExperimentGrid()))
end
println(CTExperiments.dict_to_quarto_markdown(dict, filter_empty=false))
```

@@

>>>>

### Counterfactual Outcomes

For all of our counterfactual evaluations, we proceed as follows: for each dataset we run $J$ bootstrap rounds ("No. Runs") to account for stochasticity (@nte-eval-default); for each bootstrap round, we randomly draw factual and target pairs; then, for each model, we draw samples from the test set (with replacement) for which the model predicts the randomly chosen factual class; finally, we generate multiple counterfactuals ("No. Counterfactuals") and evaluate the outcomes (@nte-eval-default). This is in line with standard practice in the related literature on CE. For our final results presented in the main paper, we rely on held-out test sets for evaluation. For tuning purposes we rely on training and/or validation sets.

@nte-eval-default presents the default hyperparameters used during evaluation for tuning purposes. For the main results presented in the paper, we use larger evaluations, specifically:

- "No. Runs": We set the number of bootstrap rounds to $J=100$ for all datasets. 
- "No. Individuals": In each round we draw 1,250, 500 and 125 samples for synthetic datasets, real-world tabular datasets and *MNIST*, respectively, across five different values for the strength of the energy penalty of *ECCCo* at test time, $\lambda_{\text{egy}}\in\{0.1, 0.5, 1.0, 5.0, 10.0\}$.

@@

dict = Logging.with_logger(Logging.NullLogger()) do  
    CTExperiments.to_dict(EvaluationGrid(ExperimentGrid()))
end
println(CTExperiments.dict_to_quarto_markdown(dict["counterfactual_params"], filter_empty=false))
```

@@

### Predictive Performance

To assess (robust) predictive performance, we evaluate model accuracy on (adversarially perturbed) test data. To generate adversarial examples we use the Fast Gradient Sign Method (FGSM) [@goodfellow2014explaining]. For the main results in the paper, we choose a range of values $\epsilon=[0.0,0.1]$. In some places of this appendix, you will also find predictive performance evaluations in terms of the F1-score.
